---
title: "Spanish Translation Statistics"
output:
  html_notebook:
    code_folding: hide
    toc_depth: 4
    toc: true
    toc_float:
      collapsed: false
---


```{r setup, results='hide', message=FALSE, warning=FALSE}
box::use(
  dplyr, ggplot2, googledrive, googlesheets4, here, purrr, readr, stringdist,
  stringr, tibble, tidyr, treemap,
  ../mod
)
```

```{r auth, include = FALSE}
googledrive$drive_auth(email = "allenbaron@som.umaryland.edu")
googlesheets4$gs4_auth(email = "allenbaron@som.umaryland.edu")
```


```{r custom_functions}
# identifies the primary sheet for each datatype (only for non-physical disorder)
which_sheet <- function(id, datatype) {
  box::use(googlesheets4, stringr)

  sheets <- googlesheets4$sheet_names(id)
  sh_regex <- paste0("^[^_0-9]+[0-9]+.*", datatype, "$")
  sh_possible <- sheets[stringr$str_detect(sheets, sh_regex)]
  sh_n <- as.numeric(stringr$str_match(sh_possible, "[0-9]+"))
  sh_final <- sh_possible[sh_n == max(sh_n)]
  if (length(sh_final) != 1) {
    stop("Could not identify primary sheet for ", datatype)
  }
  sh_final
}

# standardize data or reverse standardization (including unmerge, if necessary)
std_switch <- function(.df) {
  box::use(dplyr, purrr, stringr)

  col_en_es <- c(
    "label" = "etiqueta", "definition" = "definición", "synonym" = "sinónimo"
  )
  extra_col <- c("passed", "final")

  # standardize
  if (any(col_en_es %in% names(.df))) {
    out <- .df
    colnm <- col_en_es[col_en_es %in% names(out)]
    colnm_n <- length(colnm)
    if (colnm_n != 1) {
      stop(paste0("Exact 1 data type should be identified, not ", colnm_n))
    }

    if (!"datatype" %in% names(.df)) {
      out <-  out |>
        dplyr$mutate(
          datatype = factor(
            names(colnm),
            levels = c("label", "synonym", "definition")
          ),
          .before = 1
        )
    }

    renm_vctr <- c("source_text" = names(colnm), "target_text" = unname(colnm))
    out <- out |>
      dplyr$select(-"clase") |>
      dplyr$rename(
        "source_id" = "class",
        dplyr$all_of(renm_vctr)
      ) |>
      dplyr$rename_with(
        .fn = ~ stringr$str_remove(.x, paste0(colnm, "_"))
      )
  } else { # reverse standardization
    .df1 <- .df |>
      dplyr$rename("class" = "source_id") |>
      dplyr$mutate(clase = class, .after = "google_translate_to_en") |>
      dplyr$arrange(.data$datatype)
    df_list <- split(.df1, .df1$datatype)
    # drop empty (not all datatypes present since datatype is a factor)
    df_list <- df_list[purrr$map_lgl(df_list, ~ nrow(.x) > 0)]
    names(df_list) <- purrr$map_chr(df_list, ~ as.character(unique(.x$datatype)))

    out <- purrr$map(
      df_list,
      function(.x) {
        colnm <- col_en_es[names(col_en_es) == unique(.x$datatype)]
        renm_vctr <- purrr$set_names(
          c("source_text", "target_text"),
          nm = c(names(colnm), unname(colnm))
        )
        .x |>
          dplyr$rename(dplyr$all_of(renm_vctr)) |>
          dplyr$rename_with(
            .cols = dplyr$all_of(extra_col),
            .fn = ~ paste0(colnm, "_", .x)
          ) |>
          dplyr$select(-"datatype")
      }
    )
    if (length(out) == 1) out <- out[[1]]
  }

  out
}

count_percent <- function(.df, ..., wt = NULL, sort = FALSE, nm = NULL,
                          pct_by = NULL, digits = 2) {
  box::use(dplyr, rlang)
  if (is.null(nm)) nm <- "n"
  if (!is.null(pct_by)) {
    pct_nm <- paste0("pct_", paste0(pct_by, collapse = "_"))
  } else {
    pct_nm <- "pct"
  }
  .df |>
    dplyr$count(..., wt = !!rlang$enquo(wt), sort = sort, name = nm) |>
    dplyr$mutate(
      {{ pct_nm }} := round(100 * .data[[nm]] / sum(.data[[nm]]), digits),
      .by = dplyr$all_of(pct_by)
    )# |>
    # dplyr$mutate({{ nm }} := format(.data[[nm]], big.mark = ","))
}

tmap <- function(.df, title) {
  box::use(treemap)
  .df |>
    dplyr$filter(.data$datatype == title) |>
    treemap$treemap(
      index = c("set_overall", "details"),
      vSize = "n",
      type = "index",
      title = title
    )
}

collapse_col <- function (df, .cols, delim = "|", na.rm = FALSE) {
  box::use(dplyr)
  uniq_to_string <- function(x, delim, na.rm) {
    if (na.rm) x <- na.omit(x)
    paste0(unique(x), collapse = delim)
  }
    df |>
      dplyr$group_by(dplyr$across(-{{ .cols }})) |>
      dplyr$summarize(
        dplyr$across(
          .cols = dplyr$everything(),
          .fns = ~ uniq_to_string(.x, delim, na.rm)
        ),
        .groups = "drop"
      ) |>
      dplyr$select(dplyr$one_of(names(df)))
}

filter_id_pred <- function(df, id, pred) {
  nm_pred <- stringr$str_detect(
    pred,
    stringr$regex("label|synonym", ignore_case = TRUE)
  )
  if (nm_pred) {
    dplyr$filter(
      df,
      .data$source_id == id,
      stringr$str_detect(
        .data$predicate,
        stringr$regex("label|synonym", ignore_case = TRUE)
      )
    )
  } else {
    dplyr$filter(
      df,
      .data$source_id == id,
      .data$predicate == pred
    )
  }
}

index_id_pred <- function(df, id, pred) {
  nm_pred <- stringr$str_detect(
    pred,
    stringr$regex("label|synonym", ignore_case = TRUE)
  )
  if (nm_pred) {
    pred_match <- stringr$str_detect(
        df$predicate,
        stringr$regex("label|synonym", ignore_case = TRUE)
      )
  } else {
    pred_match <- df$predicate == pred
  }

  which(df$source_id == id & pred_match)
}

match_id_pred_fuzzy <- function(x_df, y_df,
                                include = c("predicate", "source_text")) {
  stopifnot("`id` must be same length as `x`" = length(id) == length(x))

  box::use(dplyr, purrr, stringdist, stringr)

  y_sets <- purrr$map2(x_df$id, x_df$pred, ~ index_id_pred(y_df, .x, .y))
  y_cols <- dplyr$select(y_df, dplyr$all_of(include))
  y_na <- y_cols[1, ] |>
    dplyr$mutate(dplyr$across(dplyr$everything(), ~ NA_character_))
  y_bind <- purrr$map2_chr(
    x,
    y_sets,
    function(.x, .i) {
      if (length(.i) == 0) return(y_na)
      if (length(.i) == 1) return(y_cols[.i, ])
      best_pos <- stringdist$amatch(.x, .i, maxDist = 3)
      .i[best_pos]
    }
  )
}
```


# Data Load & Preprocessing

```{r gs_info}
# all data but physical disorder
no_pd_dir <- "https://drive.google.com/drive/folders/1GqYzQ6ZI5I3pFC82XLKG5P61STxqwn4b/"

gs_no_pd <- googledrive$drive_ls(no_pd_dir, type = "spreadsheet") |>
  dplyr$mutate(
    datatype = factor(
      stringr$str_remove(name, ".*_"),
      levels = c("label", "synonym", "definition")
    ),
    batch = "remaining"
  ) |>
  dplyr$arrange(.data$datatype) |>
  dplyr$rowwise() |>
  dplyr$mutate(sheet = which_sheet(.data$id, .data$datatype))


# physical disorder
gs_pd <- googledrive$as_dribble(
  "https://docs.google.com/spreadsheets/d/1Fvnmz_3KNXuLvLtJGm9eKkqvV33mIZUESsPWv28uUAg/edit?gid=115632665#gid=115632665"
) |>
  dplyr$mutate(
    batch = "physical disorder",
    sheets = list(googlesheets4$sheet_names(.data$id))
  ) |>
  dplyr$cross_join(
    tibble$tibble(datatype = gs_no_pd$datatype)
  ) |>
  dplyr$rowwise() |>
  dplyr$mutate(
    sheet = which_sheet(.data$id, .data$datatype),
    sheets = NULL
  )

gs_all <- dplyr$bind_rows(gs_no_pd, gs_pd) |>
  dplyr$mutate(
    data = list(
      googlesheets4$read_sheet(ss = .data$id, sheet = .data$sheet)
    )
  )
```

```{r data}
all_df <- gs_all |>
  dplyr$rowwise() |>
  dplyr$select("batch", "data") |>
  dplyr$mutate(data = list(std_switch(.data$data))) |>
  tidyr$unnest("data") |>
  dplyr$mutate(
    source_id = mod$general$uri_curie(.data$source_id, to = "curie"),
    dplyr$across(
      dplyr$starts_with("match_rating"),
      .names = "{stringr$str_replace(.col, 'match_rating_(.+)', 'set_\\\\1')}",
      ~ dplyr$case_when(
        .x == 1 ~ "exact",
        .x > 0.75 ~ "passed",
        TRUE ~ "failed"
      ) |>
        factor(levels = c("exact", "passed", "failed"))
    )
  ) |>
  # drop 5 exact duplicates each from remaining labels & definitions
  unique() |>
  dplyr$mutate(
    details = dplyr$case_when(
      .data$set_en == .data$set_es ~ .data$set_overall,
      TRUE ~ paste0("en: ", .data$set_en, "\nes: ", .data$set_es)
    )
  )
```

Checking for duplicates...
```{r duplicates}
dup_df <- all_df |>
  dplyr$mutate(
    dup_chr = paste0(.data$datatype, .data$source_id, .data$source_text),
    en_dup = duplicated(.data$dup_chr) |
      duplicated(.data$dup_chr, fromLast = TRUE),
    dup_chr = NULL
  ) |>
  dplyr$filter(en_dup)

if (nrow(dup_df) > 0) {
  message("Duplicates exist from input:")
  dup_df
} else {
  message("No duplicates found.")
}
```

IDs in both sets ('remaining' and 'physical disorder')...
_checking to be sure there's not an overlap_
```{r}
both_sets <- all_df |>
  dplyr$group_by(.data$source_id) |>
  dplyr$summarize(
    in_set = dplyr$if_else(
      dplyr$n_distinct(.data$batch) == 2,
      "both",
      unique(.data$batch)
    )
  ) |>
  dplyr$filter(in_set == "both")

if (nrow(both_sets) > 0) {
  message("Some IDs exist in both sets:")
  both_sets
} else {
  message("No IDs found in both sets.")
}
```

# Statistics

## Overall data counts

### Total Translations

**Terms translated:** `r format(dplyr$n_distinct(all_df$source_id), big.mark = ",")`

```{r}
dplyr$count(all_df, .data$datatype)
```

_* **Note:**_ Some text data was not translated by TSG in the "remaining" disease set for all 3 datatypes (label, synonym, definition). This explains the discrepancy between term count and label translations.

### Consensus Counts

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_overall,
  pct_by = "datatype"
)
```


## Detailed Counts

### By Batch

#### Overall Counts

```{r}
dplyr$count(all_df, .data$batch, .data$datatype)
```


#### Consensus Counts

```{r}
all_df |>
  split(all_df$batch) |>
  purrr$map(
    ~ count_percent(
      .x,
      .data$datatype,
      .data$set_overall,
      pct_by = "datatype"
    )
  )
```

### Overall within Language Comparison Counts

#### English Only

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_en,
  pct_by = "datatype"
)
```

#### Spanish Only

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_es,
  pct_by = "datatype"
)
```

#### Overall, English & Spanish

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_overall,
  .data$set_en,
  .data$set_es,
  pct_by = c("datatype", "set_overall")
)
```

**Plots**

```{r}
detailed_sets <- count_percent(
  all_df,
  .data$datatype,
  .data$set_overall,
  .data$details,
  pct_by = c("datatype", "set_overall")
)

purrr$walk(
  c("label", "synonym", "definition"),
  ~ tmap(detailed_sets, .x)
)
```

## Text Changes

Comparing changes to release v2025-03-31 to calculate how much text has changed since data was first submitted for translation:

```{r}
pd_202310 <- mod$robot$get_obo_text(
  here$here("data/physical_disorder/doid-v2023-10-21.owl")
) |>
  dplyr$semi_join(
    dplyr$filter(all_df, batch == "physical disorder"),
    by = "source_id"
  ) |>
  dplyr$mutate(batch = "physical disorder")

remain_202402 <- mod$robot$get_obo_text(
  here$here("data/remaining_diseases/doid-v2024-02-28.owl")
) |>
  dplyr$semi_join(
    dplyr$filter(all_df, batch == "remaining"),
    by = "source_id"
  ) |>
  dplyr$mutate(batch = "remaining")

# shows how much we don't yet have as changed... probably not the right approach
# es_tsv <- readr$read_tsv(
#   here$here("data/v2025-03-31/doid-es.tsv"),
#   show_col_types = FALSE
# )
#
# es_simple <- es_tsv |>
#   dplyr$select("source_id", "predicate", "source_text") |>
#   collapse_col(-"source_text", na.rm = TRUE)


doid_202503 <- mod$robot$get_obo_text(
  here$here("data/v2025-03-31/doid.owl")
) |>
  dplyr$mutate(batch = "latest") |>
  tidyr$replace_na(list(deprecated = FALSE))

batch_date <- c(
  "physical disorder" = "2023-10-21",
  "remaining" = "2024-02-28",
  "latest" = "2025-03-31"
) |>
  as.Date()
```

```{r include = FALSE}
# checking if any text duplicates exist in latest release --> would be errors
latest_err <- doid_202503 |>
  dplyr$filter(
    !.data$deprecated,
    is.na(.data$synonym_type),
    !stringr$str_detect(.data$predicate, "Related|Broad|Narrow")
  ) |>
  collapse_col(-"source_text", delim = "|") |>
  dplyr$filter(
    dplyr$if_any(
      .cols = dplyr$everything(),
      .fns = ~ stringr$str_detect(.x, stringr$coll("|"))
    )
  ) |>
  tidyr$separate_longer_delim(
    cols = -"source_text",
    delim = "|"
  ) |>
  dplyr$mutate(dplyr$across(dplyr$everything(), ~ dplyr$na_if(.x, "NA")))

latest_err
```

```{r}
doid_simple <- dplyr$bind_rows(pd_202310, remain_202402) |>
  dplyr$arrange(.data$source_id, .data$predicate, .data$source_text) |>
  collapse_col(-"source_text", na.rm = TRUE) |>
  dplyr$select("source_id", "predicate", "source_text", "batch")

latest_simple <- doid_202503 |>
  dplyr$arrange(.data$source_id, .data$predicate, .data$source_text) |>
  collapse_col(-c("source_text", "deprecated"), na.rm = TRUE) |>
  dplyr$mutate(dplyr$across(-"deprecated", ~ dplyr$na_if(.x, "")))
```

Changes in latest compared to original sources of translation
```{r}
new <- latest_simple |>
  dplyr$filter(!.data$deprecated) |>
  dplyr$anti_join(doid_simple, by = "source_text") |>
  dplyr$mutate(change = "new") |>
  dplyr$rename_with(
    .cols = -c("source_text", "deprecated"),
    .fn = ~ paste0(.x, "_new")
  )

text_compare <- latest_simple |>
  dplyr$select(-"batch") |>
  dplyr$rename_with(
    .cols = -c("source_text", "deprecated"),
    .fn = ~ paste0(.x, "_new")
  ) |>
  dplyr$right_join(doid_simple, by = "source_text") |>
  dplyr$bind_rows(new) |>
  dplyr$mutate(
    change = dplyr$case_when(
      is.na(.data$source_id) ~ "new",
      .data$deprecated ~ "term deprecated",
      is.na(.data$predicate_new) ~ "text differs",
      .data$predicate == .data$predicate_new ~ "none",
      .data$predicate != .data$predicate_new &
        .data$source_id == .data$source_id_new ~ "predicate differs",
      .data$predicate != .data$predicate_new ~ "source_id & predicate differ"
    )
  )

# too slow to execute
# text_differ <- text_compare |>
#   dplyr::filter(change == "text differs") |>
#   dplyr$mutate(
#     new_text_pos = stringdist$amatch(
#       .data$source_text,
#       new$source_text, maxDist = 3)
#   )

count_percent(text_compare, .data$change)
```

**New by predicate:**

```{r}
count_percent(new, .data$predicate_new)
```



The specific text changes are as follows:
```{r}
text_changes <- text_compare |>
  dplyr$filter(change == "text differs") |>
  dplyr$select("source_id", "predicate", "source_text", "batch") |>
  dplyr$mutate(
    new_text = match_id_pred_fuzzy(
      .data$source_id,
      .data$predicate,
      .data$source_text,
      doid_202503
    )
  )

text_changes
```


## Latest Spanish Coverage (v2025-03-31)

```{r}
di <- mod$robot$get_obo_text(
  here$here("data/v2025-03-31/doid-international.owl")
) |>
  dplyr$filter(
    stringr$str_detect(.data$source_id, stringr$coll("DOID")),
    is.na(.data$synonym_type),
    is.na(.data$deprecated)
  )

dplyr$count(di, .data$source_lang) |>
  tidyr$pivot_wider(
    names_from = "source_lang",
    values_from = "n"
  ) |>
  dplyr$mutate(pct_coverage = round(.data$es / .data$en * 100, 2))
```

**By predicate:**

```{r}
dplyr$count(di, .data$source_lang, .data$predicate) |>
  tidyr$pivot_wider(
    names_from = "source_lang",
    values_from = "n"
  ) |>
  dplyr$mutate(pct_coverage = round(.data$es / .data$en * 100, 2))
```
