---
title: "Spanish Translation Statistics"
output: 
  html_notebook:
    code_folding: hide
    toc_depth: 4
    toc: true
    toc_float: true
      collapsed: false
---

```{r auth, include = FALSE}
googledrive::drive_auth(email = "allenbaron@som.umaryland.edu")
googlesheets4::gs4_auth(email = "allenbaron@som.umaryland.edu")
```

```{r setup, results='hide', message=FALSE, warning=FALSE}
box::use(
  dplyr, ggplot2, googledrive, googlesheets4, purrr, stringr, tibble, tidyr, treemap,
  ../mod
)
```

```{r custom_functions}
# identifies the primary sheet for each datatype (only for non-physical disorder)
which_sheet <- function(id, datatype) {
  box::use(googlesheets4, stringr)

  sheets <- googlesheets4$sheet_names(id)
  sh_regex <- paste0("^[^_0-9]+[0-9]+.*", datatype, "$")
  sh_possible <- sheets[stringr$str_detect(sheets, sh_regex)]
  sh_n <- as.numeric(stringr$str_match(sh_possible, "[0-9]+"))
  sh_final <- sh_possible[sh_n == max(sh_n)]
  if (length(sh_final) != 1) {
    stop("Could not identify primary sheet for ", datatype)
  }
  sh_final
}

# standardize data or reverse standardization (including unmerge, if necessary)
std_switch <- function(.df) {
  box::use(dplyr, purrr, stringr)

  col_en_es <- c(
    "label" = "etiqueta", "definition" = "definición", "synonym" = "sinónimo"
  )
  extra_col <- c("passed", "final")

  # standardize
  if (any(col_en_es %in% names(.df))) {
    out <- .df
    colnm <- col_en_es[col_en_es %in% names(out)]
    colnm_n <- length(colnm)
    if (colnm_n != 1) {
      stop(paste0("Exact 1 data type should be identified, not ", colnm_n))
    }
    
    if (!"datatype" %in% names(.df)) {
      out <-  out |>
        dplyr$mutate(
          datatype = factor(
            names(colnm),
            levels = c("label", "synonym", "definition")
          ),
          .before = 1
        )
    }
    
    renm_vctr <- c("source_text" = names(colnm), "target_text" = unname(colnm))
    out <- out |>
      dplyr$select(-"clase") |>
      dplyr$rename(
        "source_id" = "class",
        dplyr$all_of(renm_vctr)
      ) |>
      dplyr$rename_with(
        .fn = ~ stringr::str_remove(.x, paste0(colnm, "_"))
      )
  } else { # reverse standardization
    .df1 <- .df |>
      dplyr$rename("class" = "source_id") |>
      dplyr$mutate(clase = class, .after = "google_translate_to_en") |>
      dplyr$arrange(.data$datatype)
    df_list <- split(.df1, .df1$datatype)
    # drop empty (not all datatypes present since datatype is a factor)
    df_list <- df_list[purrr::map_lgl(df_list, ~ nrow(.x) > 0)]
    names(df_list) <- purrr$map_chr(df_list, ~ as.character(unique(.x$datatype)))
    
    out <- purrr::map(
      df_list,
      function(.x) {
        colnm <- col_en_es[names(col_en_es) == unique(.x$datatype)]
        renm_vctr <- purrr$set_names(
          c("source_text", "target_text"),
          nm = c(names(colnm), unname(colnm))
        )
        .x |>
          dplyr$rename(dplyr$all_of(renm_vctr)) |>
          dplyr$rename_with(
            .cols = dplyr$all_of(extra_col),
            .fn = ~ paste0(colnm, "_", .x)
          ) |>
          dplyr$select(-"datatype")
      }
    )
    if (length(out) == 1) out <- out[[1]]
  }

  out
}

count_percent <- function(.df, ..., wt = NULL, sort = FALSE, nm = NULL,
                          pct_by = NULL, digits = 2) {
  box::use(dplyr, rlang)
  if (is.null(nm)) nm <- "n"
  if (!is.null(pct_by)) {
    pct_nm <- paste0("pct_", paste0(pct_by, collapse = "_"))
  } else {
    pct_nm <- "pct"
  }
  .df |>
    dplyr$count(..., wt = !!rlang$enquo(wt), sort = sort, name = nm) |>
    dplyr$mutate(
      {{ pct_nm }} := round(100 * .data[[nm]] / sum(.data[[nm]]), digits),
      .by = dplyr$all_of(pct_by)
    )# |>
    # dplyr$mutate({{ nm }} := format(.data[[nm]], big.mark = ","))
}

tmap <- function(.df, title) {
  box::use(treemap)
  .df |> 
    dplyr::filter(.data$datatype == title) |>
    treemap$treemap(
      index = c("set_overall", "details"),
      vSize = "n",
      type = "index",
      title = title
    )
}
```


# Data Load & Preprocessing

```{r gs_info}
# all data but physical disorder 
no_pd_dir <- "https://drive.google.com/drive/folders/1GqYzQ6ZI5I3pFC82XLKG5P61STxqwn4b/"

gs_no_pd <- googledrive$drive_ls(no_pd_dir, type = "spreadsheet") |>
  dplyr$mutate(
    datatype = factor(
      stringr$str_remove(name, ".*_"),
      levels = c("label", "synonym", "definition")
    ),
    batch = "remaining"
  ) |>
  dplyr$arrange(.data$datatype) |>
  dplyr$rowwise() |>
  dplyr$mutate(sheet = which_sheet(.data$id, .data$datatype))


# physical disorder
gs_pd <- googledrive$as_dribble(
  "https://docs.google.com/spreadsheets/d/1Fvnmz_3KNXuLvLtJGm9eKkqvV33mIZUESsPWv28uUAg/edit?gid=115632665#gid=115632665"
) |>
  dplyr$mutate(
    batch = "physical disorder",
    sheets = list(googlesheets4$sheet_names(.data$id))
  ) |>
  dplyr$cross_join(
    tibble$tibble(datatype = gs_no_pd$datatype)
  ) |>
  dplyr$rowwise() |>
  dplyr$mutate(
    sheet = which_sheet(.data$id, .data$datatype),
    sheets = NULL
  )

gs_all <- dplyr$bind_rows(gs_no_pd, gs_pd) |>
  dplyr$mutate(
    data = list(
      googlesheets4$read_sheet(ss = .data$id, sheet = .data$sheet)
    )
  )
```

```{r data}
all_df <- gs_all |>
  dplyr$rowwise() |>
  dplyr$select("batch", "data") |>
  dplyr$mutate(data = list(std_switch(.data$data))) |>
  tidyr$unnest("data") |>
  dplyr$mutate(
    source_id = mod$general$uri_curie(.data$source_id, to = "curie"),
    dplyr::across(
      dplyr::starts_with("match_rating"),
      .names = "{stringr$str_replace(.col, 'match_rating_(.+)', 'set_\\\\1')}",
      ~ dplyr::case_when(
        .x == 1 ~ "exact",
        .x > 0.75 ~ "passed",
        TRUE ~ "failed"
      ) |>
        factor(levels = c("exact", "passed", "failed"))
    )
  ) |>
  # drop 5 exact duplicates each from remaining labels & definitions
  unique() |>
  dplyr::mutate(
    details = dplyr$case_when(
      .data$set_en == .data$set_es ~ .data$set_overall,
      TRUE ~ paste0("en: ", .data$set_en, "\nes: ", .data$set_es)
    )
  )
```

Checking for duplicates...
```{r duplicates}
dup_df <- all_df |>
  dplyr$mutate(
    dup_chr = paste0(.data$datatype, .data$source_id, .data$source_text),
    en_dup = duplicated(.data$dup_chr) |
      duplicated(.data$dup_chr, fromLast = TRUE),
    dup_chr = NULL
  ) |>
  dplyr::filter(en_dup)

if (nrow(dup_df) > 0) {
  message("Duplicates exist from input:")
  dup_df
} else {
  message("No duplicates found.")
}
```

IDs in both sets ('remaining' and 'physical disorder')...
_checking to be sure there's not an overlap_
```{r}
both_sets <- all_df |>
  dplyr$group_by(.data$source_id) |>
  dplyr$summarize(
    in_set = dplyr$if_else(
      dplyr::n_distinct(.data$batch) == 2,
      "both",
      unique(.data$batch)
    )
  ) |>
  dplyr$filter(in_set == "both")

if (nrow(both_sets) > 0) {
  message("Some IDs exist in both sets:")
  both_sets
} else {
  message("No IDs found in both sets.")
}
```

# Statistics

## Overall data counts

### Total Translations

**Terms translated:** `r format(dplyr::n_distinct(all_df$source_id), big.mark = ",")`

```{r}
dplyr$count(all_df, .data$datatype)
```

_* **Note:**_ Some text data was not translated by TSG in the "remaining" disease set for all 3 datatypes (label, synonym, definition). This explains the discrepancy between term count and label translations.

### Consensus Counts

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_overall,
  pct_by = "datatype"
)
```


## Detailed Counts

### By Batch

#### Overall Counts

```{r}
dplyr$count(all_df, .data$batch, .data$datatype)
```


#### Consensus Counts

```{r}
all_df |>
  split(all_df$batch) |>
  purrr::map(
    ~ count_percent(
      .x,
      .data$datatype,
      .data$set_overall,
      pct_by = "datatype"
    )
  )
```

### Overall within Language Comparison Counts

#### English Only

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_en,
  pct_by = "datatype"
)
```

#### Spanish Only

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_es,
  pct_by = "datatype"
)
```

#### Overall, English & Spanish

```{r}
count_percent(
  all_df,
  .data$datatype,
  .data$set_overall,
  .data$set_en,
  .data$set_es,
  pct_by = c("datatype", "set_overall")
)
```

**Plots**

```{r}
detailed_sets <- count_percent(
  all_df,
  .data$datatype,
  .data$set_overall,
  .data$details,
  pct_by = c("datatype", "set_overall")
)

purrr::walk(
  c("label", "synonym", "definition"),
  ~ tmap(detailed_sets, .x)
)
```
